Windowed Residual-Minimization Model Reduction for Dynamical Systems

### 2019-10-02 main.tex
eric latest

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach minimizes the
	time-continuous full-order-model residual over a low-dimensional 
	trial subspace across a sequence of time windows and addresses key deficiencies in existing model-reduction approaches; e.g., the time step and time scheme dependence of least-squares Petrov--Galerkin (LSPG), the exponentially growing error bounds present in Galerkin and LSPG, and the offline cost of space--time LSPG. 
We consider two types of trial subspaces
	within the proposed framework:
	one that reduces the spatial dimension of the full-order model, and one that
	reduces both the spatial and temporal dimensions of the full-order model. For both types of trial subspaces, we consider two differing solution techniques: direct (i.e., discretize-then-optimize) and indirect (i.e., optimize-then-discretize). Critically, we show that various instances of this framework yield existing model-reduction approaches, namely Galerkin, Least-Squares Petrov–Galerkin, and Space–Time Least-Squares Petrov– Galerkin. Numerical experiments on the compressible Euler and Navier--Stokes equations demonstrate that, at an increased computational cost, the \methodAcronym\ formulation can yield more accurate and physically relevant solutions than the Galerkin and Least-Squares Petrov--Galerkin approaches.

@@

Simulating parameterized dynamical systems underpins many applications across
science and engineering, including engineering design, control, financial
analysis, and epidemiology. In many contexts, executing a dynamical-system
simulation at a single parameter instance---which entails the numerical
integration of a system of ordinary differential equations (ODEs)---incurs an
extremely large computational expense.  This occurs, for example, when the
state-space dimension is large (e.g., due to fine spatial resolution when
discretizing a partial-differential-equation) and/or when the number
of time instances is large (e.g., due to stability or accuracy requirements
that limit the time step).  When the application is time-critical or
many-query in nature, analysts must replace such large-scale parameterized
dynamical-system models with a low-cost approximation that makes the
application tractable.

@@

Projection-based reduced-order models (ROMs) comprise one such approximation
strategy. First, these techniques execute a computationally expensive
\textit{offline} stage that computes a low-dimensional ``trial'' subspace on
which the (parameterized) dynamical-system state can be well approximated
(e.g., by computing state ``snapshots'' at different parameter instances, by
solving Lyapunov equations). Second, these methods execute an inexpensive
\textit{online} stage during which they compute approximations to the
dynamical-system trajectory that reside on this trial subspace via a
residual-minimization or projection process.

Model reduction for linear-time-invariant systems (and other strongly
structured dynamical systems) is quite mature \cite{wilcox_benner_rev,moore,roberts,GugercinIRKA} as system-theoretic properties (e.g., controllability,
observability, asymptotic stability, $\mathcal H_2$-optimality) can be readily
quantified and accounted for in the projection process, which yields
reduced-order models that inherit such properties.  The primary challenge for
developing reduced-order models for general nonlinear dynamical systems is
that such properties are difficult to assess. As a result, it
is difficult to develop reduced-order models that preserve important
dynamical-system properties, a fact that can lead to inaccurate or unstable methods.  To address this, researchers have
pursued several directions that aim to imbue reduced-order models of nonlinear
dynamical systems with properties that can improve robustness and accuracy.
These
efforts include residual-minimizing
projection that imbues the ROM solution with a notion of optimality~\cite{l1,carlberg_lspg,carlberg_gnat,legresley_1,legresley_2,legresley_3,bui_resmin_steady,bui_unsteady,rovas_thesis,carlberg_thesis,bui_thesis};
space--time
projection that leads to error bounds that grow slowly in time~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st};
``energy-based'' inner
products that ensure non-increasing entropy in the ROM solution~\cite{rowley_pod_energyproj,Kalashnikova_sand2014};
basis-adaptation methods that improve the ROM's accuracy \textit{a posteriori}~\cite{carlberg_hadaptation,adeim_peherstorfer}, stabilizing subspace rotations that account for truncated modes \textit{a priori}~\cite{basis_rotation}, structure-preserving
methods that enforce conservation \cite{carlberg_conservative_rom} or
(port-)Hamiltonian/Lagrangian structure
\cite{carlberg2012spd,structurePreserveBeattie,chaturantabut2016structure,farhat2014dimensional,LALL2003304} in the ROM; and subgrid-scale
modeling methods that aim to improve accuracy by addressing the closure
problem~\cite{san_iliescu_geostrophic,iliescu_pod_eddyviscosity,iliescu_vms_pod_ns,Bergmann_pod_vms,iliescu_ciazzo_residual_rom,Wang_ROM_thesis,wentland_apg,Wang:269133,San2018}.
We note that these techniques are often not mutually exclusive.
%Residual-minimizing projection methods are the most relevant to the current work and comprise the focus of the following review.
Residual-minimizing and space--time projection are the most relevant classes of
methods to the current work and comprise the focus of the following review.
\KTC{Shouldn't we review space-time also?}

@@

%\begin{itemize}
%\item Galerkin is inaccurate/unstable for non-symmetric static systems -- motivates residual minimization
%\item Residual minimization (stems from FEM): minimizes the residual of the full-order model over the range of the trial space -- originally developed for static systems
%\item Can guarantee stability for linear static systems 
%\item Extension to time-varying systems is challenging 
%\begin{itemize}
%\item What is the definition of the ODE residual?
%\item Over what time-interval is the residual being minmized?
%\end{itemize}
%\end{itemize} 
Least-squares residual minimization as a numerical solution technique for (partial) differential equations was pioneered in the finite element community and led to what is known as the least-squares finite element method (LSFEM)~\cite{ls_review,bochev_leastsquares}. Motivated by the fact that the Galerkin method can be inaccurate and unstable for nonlinear and non-self-adjoint systems, the LSFEM involves (1.) forming a quadratic functional comprising the residual norm of the partial differential equation of interest and (2.) leveraging variational principles to derive the stationary conditions of the quadratic functional; i.e, LSFEM can be thought of as finding the solution within a finite element trial space the minimizes the $2$-norm of the PDE residual. Least-squares methods often exhibit numerous advantages (e.g., least-squares system are symmetric and positive definite) over the more standard Galerkin discretization.

Residual minimization methods were extended-to and developed-for reduced-order models by a variety of researchers. Similar to LSFEM, residual minimization approaches for reduced-order models comprise finding the solution within a low-dimensional trial subspace that minimizes the residual norm of the full-order model. Residual minimization approaches were first applied to static (i.e., no time-dependence) reduced-order models in~\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis} and then extended to dynamical systems in~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}. Specifically, Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg} formulated the residual minimization problem for dynamical systems by sequentially minimizing the \textit{fully discrete} residual (i.e., the residual after temporal discretization) at each time-step instance. This formulation is most commonly known as the \textit{Least--Squares Petrov--Galerkin} (LSPG) method. Ref.~\cite{carlberg_lspg_v_galerkin} provided numerous analyses on this formulation and examined connections with the standard Galerkin method and Petrov--Galerkin methods. Critically, it was shown that (1.) for various time-stepping methods LSPG has an equivalent continuous form that corresponds to a Petrov--Galerkin projection and (2.) the standard Galerkin method minimizes the time-instantaneous residual of the full-order model. Numerous numerical experiments have demonstrated that LSPG yields more stable (although not always more accurate) solutions than the Galerkin method~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}. The common physical explanation for this improved performance is that, by minimizing the fully discrete residual, LSPG minimizes the residual over a \textit{finite-sized time-window}, rather than instantaneously as is done in the Galerkin method.

The LSPG approach only guarantees residual minimization over a single time-step, and as such is not associated with residual minimization of the full dynamical system.  
Indeed, LSPG is not equipped with \textit{a priori} guarantees of stability or convergence, even for linear time-invariant systems~\cite{bui_thesis}. Further, as it is defined at the fully discrete level, the performance of LSPG is subject to the time-step size: changing the time-step size changes the window over which the residual is minimized \textit{as well as} the temporal discretization error. As a result, LSPG provides optimal results at an intermediary time-step size. For too small a time-step, LSPG yields results similar to the Galerkin approach; in the limiting case of $\Delta t \rightarrow 0$, LSPG recovers Galerkin. For too-large a time-step, the accuracy of LSPG degrades. It is unclear if this is due to the numerical error associated with enlarging the time-step size, or rather if it is due to the size of the window the residual is being minimized over. The LSPG approach is additionally sensitive to the time discretization scheme. When used with explicit temporal discretization, for instance, LSPG again reverts to Galerkin. To gain the enhanced stability associated with LSPG, the underlying time-discretization scheme must be implicit with an appropriate time-step. This can increase the cost of the reduced-order model, especially in the case where the optimal time-step required by LSPG is small.

@@

The resulting approach is best viewed as a hybrid space--time method. A
variety of research has examined the formulation of space--time reduced-order
models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis}
and the present approach displays commonalities with several such efforts.
First,~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically the thesis
of optimize then discretize vs discretize then optimize was mentioned. This
formulation, however, was not fully developed and was eschewed for the
time-local residual minimization method (i.e., LSPG). The present work thus
formally develops (and extends) several of the concepts put forth
in~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1.) in addition to trial subspaces with space--time dimension reduction, we
consider trial subspaces with spatial dimension reduction only (i.e., a classic
spatial basis is employed, opposed to a space--time basis) -- in which case
the generalized coordinates comprise a vector of functions, and (2.) we
minimize the time-continuous residual over space--time
windows;~\cite{constantine_strom} considers the solution of a fully discrete
optimization problem over the entire space--time domain. It is noted that,
similar to the current work, a time-continuous residual-minimization
formulation is used as a starting point in~\cite{constantine_strom}, and as
such thematic similarities exist between~\cite{constantine_strom} and the
present work. Lastly, Ref.~\cite{choi_stlspg} develops a space--time
formulation of the least-squares Petrov--Galerkin approach. The formulation
minimizes the \textit{fully discrete} residual over the entire space--time
domain within a given space--time trial subspace. 
%These space--time trial spaces are constructed through higher-order singular value decompositions (SVDs), such as the truncated high-order SVD. 
The present work distinguishes itself from Ref.~\cite{choi_stlspg} in that
(1.) we perform model reduction of the time-continuous system, (2.) we
consider trial subspaces with both space--time dimension reduction and spatial
dimension reduction, and (3.) residual minimization is performed over time
windows; Ref.~\cite{choi_stlspg} considers space--time bases and minimizes the
fully-discrete residual over the entire domain.

>>>>

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach minimizes the
	time-continuous full-order-model residual over a low-dimensional trial
	subspace across a sequence of time windows and addresses key deficiencies in
	existing model-reduction approaches, e.g., the time-step and time-scheme
	dependence of least-squares Petrov--Galerkin (LSPG) projection, \textit{a
	posteriori} error bounds that grow exponentially in time for Galerkin and
	LSPG projection, and the offline cost of space--time LSPG projection.
	\KTC{How do we address the offline cost of ST-LSPG?}
We consider two types of trial subspaces
	within the proposed framework:
	one that reduces only the spatial dimension of the full-order model, and one that
	reduces both the spatial and temporal dimensions of the full-order model.
	For each type of trial subspaces, we consider two different solution
	techniques: direct (i.e., discretize-then-optimize) and indirect (i.e.,
	optimize-then-discretize). Critically, we show that various instances of
	this framework yield existing model-reduction approaches, namely Galerkin,
	LSPG, and space–time LSPG projection. Numerical experiments on the
	compressible Euler and Navier--Stokes equations demonstrate that, at an
	increased computational cost, the \methodAcronym\ formulation can yield more
	accurate and physically relevant solutions than Galerkin and
	LSPG projection.  \KTC{So which is better? Should we emphasize that it is
	more general?}

@@

Simulating parameterized dynamical systems underpins many applications across
science and engineering, including engineering design, control, financial
analysis, and epidemiology. In many contexts, executing a dynamical-system
simulation at a single parameter instance---which entails the numerical
integration of a system of ordinary differential equations (ODEs)---incurs an
extremely large computational expense.  This occurs, for example, when the
state-space dimension is large (e.g., due to fine spatial resolution when
discretizing a partial differential equation) and/or when the number of time
instances is large (e.g., due to time-step limitations incurred by stability
or accuracy considerations).  When the application is time critical or many
query in nature, analysts must replace such large-scale parameterized
dynamical-system models with a low-cost approximation that makes the
application tractable.

@@

Projection-based reduced-order models (ROMs) comprise one such approximation
strategy. First, these techniques execute a computationally expensive
\textit{offline} stage that computes a low-dimensional ``trial'' subspace on
which the dynamical-system state can be well approximated (e.g., by computing
state ``snapshots'' at different time and parameter instances, by solving
Lyapunov equations). Second, these methods execute an inexpensive
\textit{online} stage during which they compute approximations to the
dynamical-system trajectory that reside on this trial subspace via a
projection process.

Model reduction for linear-time-invariant systems (and other strongly
structured dynamical systems) is quite mature \cite{wilcox_benner_rev,moore,roberts,GugercinIRKA} as system-theoretic properties (e.g., controllability,
observability, asymptotic stability, $\mathcal H_2$-optimality) can be readily
quantified and accounted for in the projection process, which yields
reduced-order models that inherit such properties.  The primary challenge for
developing reduced-order models for general nonlinear dynamical systems is
that such properties are difficult to quantitatively assess. As a result, it
is difficult to develop reduced-order models that preserve important
dynamical-system properties, which often results in inaccurate or unstable
methods.  To address this, researchers have pursued several directions that
aim to imbue reduced-order models for nonlinear
dynamical systems with properties that can improve robustness and accuracy.
These
efforts include residual-minimizing
projection that equips the ROM solution with a notion of optimality~\cite{l1,carlberg_lspg,carlberg_gnat,legresley_1,legresley_2,legresley_3,bui_resmin_steady,bui_unsteady,rovas_thesis,carlberg_thesis,bui_thesis};
space--time
projection that leads to error bounds that grow slowly in time~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,benner_st};
``energy-based'' inner
products that ensure non-increasing entropy in the ROM solution~\cite{rowley_pod_energyproj,Kalashnikova_sand2014};
basis-adaptation methods that improve the ROM's accuracy \textit{a
posteriori}~\cite{carlberg_hadaptation,adeim_peherstorfer,etter2019online}, stabilizing subspace rotations that account for truncated modes \textit{a priori}~\cite{basis_rotation}, structure-preserving
methods that enforce conservation \cite{carlberg_conservative_rom} or
(port-)Hamiltonian/Lagrangian structure
\cite{LALL2003304,carlberg2012spd,structurePreserveBeattie,chaturantabut2016structure,farhat2014dimensional} in the ROM; and subgrid-scale
modeling methods that aim to improve accuracy by addressing the closure
problem~\cite{san_iliescu_geostrophic,iliescu_pod_eddyviscosity,iliescu_vms_pod_ns,Bergmann_pod_vms,iliescu_ciazzo_residual_rom,Wang_ROM_thesis,wentland_apg,Wang:269133,San2018}.
We note that these techniques are often not mutually exclusive.
%Residual-minimizing projection methods are the most relevant to the current work and comprise the focus of the following review.
Residual-minimizing and space--time projection are the most relevant classes of
methods to the current work and comprise the focus of the following review.

@@

%\begin{itemize}
%\item Galerkin is inaccurate/unstable for non-symmetric static systems -- motivates residual minimization
%\item Residual minimization (stems from FEM): minimizes the residual of the full-order model over the range of the trial space -- originally developed for static systems
%\item Can guarantee stability for linear static systems 
%\item Extension to time-varying systems is challenging 
%\begin{itemize}
%\item What is the definition of the ODE residual?
%\item Over what time-interval is the residual being minmized?
%\end{itemize}
%\end{itemize}

\KTC{Why are we reviewing LSFEM? This isn't a ROM. Don't think it's relevant}
Least-squares residual minimization as a numerical solution technique for
partial differential equations was pioneered in the finite-element community
and led to what is known as the least-squares finite-element method
(LSFEM)~\cite{ls_review,bochev_leastsquares}. 
\KTC{Specifically, LSFEM was designed to solve the inf--sup stability problem.
Be more precise.}
Motivated by the fact that the
Galerkin method can be inaccurate and unstable for nonlinear and
non-self-adjoint systems, the LSFEM involves (1.) forming a quadratic
functional comprising the residual norm of the partial differential equation
of interest and (2.) leveraging variational principles to derive the
stationary conditions of the quadratic functional; i.e, LSFEM can be thought
of as finding the solution within a finite element trial space the minimizes
the $2$-norm of the PDE residual. Least-squares methods often exhibit numerous
advantages (e.g., least-squares system are symmetric and positive definite)
\KTC{This is not an `advantage'. It is guaranteed to be inf--sup stable.}
over the more standard Galerkin discretization.

Researchers also developed residual-minimization methods for model reduction.
Similar to LSFEM, residual minimization in this context associates with
computing the solution within a low-dimensional trial space that minimizes the
$\ell^2$-norm of the full-order-model residual. \KTC{haven't defined
`full-order model'} Researchers have developed such residual-minimizing
model-reduction methods for both static systems (i.e., systems without
time-dependence)
\cite{legresley_1,legresley_2,legresley_3,bui_resmin_steady,rovas_thesis,carlberg_thesis,bui_thesis}
and dynamical
systems~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg,carlberg_lspg_v_galerkin}.
In the latter category,
Refs.~\cite{bui_thesis,bui_unsteady,carlberg_thesis,carlberg_gnat,carlberg_lspg}
formulated the residual minimization problem for dynamical systems by
sequentially minimizing the \textit{time-discrete} full-order-model residual
(i.e., the residual arising after applying time discretization) at each time
instance on the time-discretization grid. This formulation is often referred
to as the \textit{least--squares Petrov--Galerkin} (LSPG) method.
Ref.~\cite{carlberg_lspg_v_galerkin} performed detailed analyses of this
formulation and examined its connections with Galerkin projection. Critically,
this work demonstrated that (1) under certain conditions, LSPG can be derived
from a Petrov--Galerkin projection executed on the time-continuous
full-order-model residual, and (2) LSPG and Galekrin projection are equivalent
in the limit as the time step goes to zero (i.e., Galerkin projection
minimizes the time-instantaneous full-order model residual). Numerous
numerical experiments have demonstrated that LSPG projection often yields more
accurate and stable solutions than Galerkin
projection~\cite{bui_thesis,carlberg_lspg_v_galerkin,carlberg_gnat,carlberg_thesis,parish_apg}.
The common intuitive explanation for this improved performance is that, by
minimizing the full-order-model residual over a finite time window (rather
than time instantanously), LSPG projection computes solutions that are more
accurate over a larger part the trajectory compared with Galerkin projection.

The LSPG approach sequentially performs residual minimization over a single
time-step, and as such is not associated with residual minimization over the
entire time interval. \KTC{This is what ST-LSPG does. Why do we ignore it? the
full dynamical system} Indeed, LSPG is not equipped with \textit{a priori}
guarantees of stability or convergence, even for linear time-invariant
systems~\cite{bui_thesis}. Further, LSPG performs residual minimization over a
time step, its performance is affected by the time-step size in two distinct
ways: changing the time-step size modifies the window over which the residual
is minimized \textit{as well as} the time-discretization error. As a result,
LSPG often yields the lowest error for an intermediate value of the time step.
For too small a time-step, LSPG yields results similar to the Galerkin
approach; in the limiting case of $\Delta t \rightarrow 0$, LSPG recovers
Galerkin. For too-large a time-step, the accuracy of LSPG degrades. It is
unclear if this is due to the numerical error associated with enlarging the
time-step size, or rather if it is due to the size of the window the residual
is being minimized over. The LSPG approach is additionally sensitive to the
time discretization scheme. When used with explicit temporal discretization,
for instance, LSPG again reverts to Galerkin. To gain the enhanced stability
associated with LSPG, the underlying time-discretization scheme must be
implicit with an appropriate time-step. This can increase the cost of the
reduced-order model, especially in the case where the optimal time-step
required by LSPG is small.

@@

The resulting approach is best viewed as a hybrid space--time method. A
variety of research has examined the formulation of space--time reduced-order
models~\cite{choi_stlspg,constantine_strom,URBAN2012203,Yano2014ASC,bui_thesis}
and the present approach displays commonalities with several such efforts.
First,~\cite{bui_thesis} briefly formulated a space--time least-squares ROM
and connected this formulation with optimal control; specifically the thesis
of optimize then discretize vs discretize then optimize was mentioned. This
formulation, however, was not fully developed and was eschewed for the
time-local residual minimization method (i.e., LSPG). The present work thus
formally develops (and extends) several of the concepts put forth
in~\cite{bui_thesis}. Next, Ref.~\cite{constantine_strom} developed a
space--time residual minimization formulation for model interpolation.  The
present work distinguishes itself from Ref.~\cite{constantine_strom} in that
(1.) in addition to trial subspaces with space--time dimension reduction, we
consider trial subspaces with spatial dimension reduction only (i.e., a classic
spatial basis is employed, opposed to a space--time basis) -- in which case
the generalized coordinates comprise a vector of functions, and (2.) we
minimize the time-continuous residual over space--time
windows;~\cite{constantine_strom} considers the solution of a fully discrete
optimization problem over the entire space--time domain. It is noted that,
similar to the current work, a time-continuous residual-minimization
formulation is used as a starting point in~\cite{constantine_strom}, and as
such thematic similarities exist between~\cite{constantine_strom} and the
present work. Lastly, Ref.~\cite{choi_stlspg} develops a space--time
formulation of the least-squares Petrov--Galerkin approach. The formulation
minimizes the \textit{fully discrete} residual over the entire space--time
domain within a given space--time trial subspace. 
%These space--time trial spaces are constructed through higher-order singular value decompositions (SVDs), such as the truncated high-order SVD. 
The present work distinguishes itself from Ref.~\cite{choi_stlspg} in that
(1.) we perform model reduction of the time-continuous system, (2.) we
consider trial subspaces with both space--time dimension reduction and spatial
dimension reduction, and (3.) residual minimization is performed over time
windows; Ref.~\cite{choi_stlspg} considers space--time bases and minimizes the
fully-discrete residual over the entire time domain.
